Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 paper.tex
Move tables around
Spell check

Conversion was performed on the All of Us Researcher Workbench platform
(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB
of RAM, which cost \$3.80/hour. We first converted the VCF to  ICF, which
took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU
time, achieving a parallel efficiency of 86\%. This step incurred a cost of
\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed
across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with
a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a
cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across
232,809 chunk files.

@@

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
whole genome alignments and metadata 
(4,384,310 samples, 29903 variants)
consisting of 37 fields and 391 MiB of storage, 
of which 280MiB is used by aligments ($\sim 81$X smaller 
than gzipped FASTA). The chunk size is 100 variants $\times$
10,000 samples.
Shown are the top 9 fields consuming at least 1\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
Field names (derived from source TSV file) are shortened for 
display purposes by replacing ``Genbank'' with ``G''
and ``Viridian'' with ``V''.
\label{tab-sc2-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
\bottomrule
\end{tabular}
\end{table}

Decompressed, the Viridian dataset consists of 48
FASTA files totaling 125 GiB of storage and a 1.4 GiB
TSV file for sample metadata.
When aligned to the reference and converted to Zarr, the
entire dataset is stored in a single 391 MiB Zip file.
When compressed with bgzip (so they can be indexed; see next 
paragraph) the alignments require 22 GiB of storage.
Table~\ref{tab-sc2-data} summarises the storage used by individual
arrays, and shows that the alignment data requires 298 MiB,
around 81X smaller than gzipped FASTA.
The original dataset is distributed as xz compressed FASTA files, which
achieves even higher compression levels at a total of 218 MiB.
However, these files must be decompressed in order to access the data
using standard FASTA processing techniques.

@@

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing
high-performance analysis tools,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. 
Using the TensorStore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend 
(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats),
for which the core functionality is defined in a 264 line C++ file.
TensorStore provides efficient read and write access to array
data in several formats, used, for example, in
checkpointing~\cite{orbax-tensorstore} 
TensorFlow model training~\citep{tensorflow2015-whitepaper}.
It supports several cloud platforms, and is based on an
asyncronous, parallel processing model (see Discussion).

@@

\subsection{Future applications: GPU acceleration}
GPU acceleration has revolutionized scientific computing by leveraging 
their massively parallel architecture to achieve significant speedups in
computationally intensive tasks such as simulations, data analysis, and machine
learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
mittal2014survey,owens2008gpu}
This has had a transformative impact across many 
% TODO More
fields~\citep[e.g.][]{stone2007accelerating,pandey2022transformational}.
While early GPU programming required expertise in
low-level languages like CUDA or OpenCL, the advent of high-level Python
libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
and PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
has made GPU acceleration far more accessible.
Deep learning~\citep{lecun2015deep} has led to major breakthroughs in 
several domains in recent 
% TODO more?
years~\citep[e.g.][]{jumper2021highly,merchant2023scaling}
and depends crucially on processing large volumes of data
on GPUs or other custom hardware accelerators.
The Python data science ecosystem is particularly powerful in this area,
with GPU-native libraries like
PyTorch~\cite{paszke2019pytorch}
and TensorFlow~\citep{tensorflow2015-whitepaper}
providing the basic infrastructure for the field.

@@

To demonstate the ease at which non-trivial, novel calculations 
can be performed on a GPU,
we first reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the central operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel. 
This is very similar to the Numba compiled function used in other
experiments, with minimal changes required to take advantage of
on-GPU parallelism. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82 MiB), decompress each to a local
NumPy array and then copy the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with a
data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460 MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer time to 138ms and processing time to 6ms.

>>>>

Conversion was performed on the All of Us Researcher Workbench platform
(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB
of RAM, which cost \$3.80/hour. We first converted the VCF to  ICF, which
took 6 hours and 17 minutes of effective runtime and used 343 hours of CPU
time, achieving a parallel efficiency of 86\%. This step incurred a cost of
\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed
across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with
a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a
cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across
232,809 chunk files.

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
whole genome alignments and metadata 
(4,384,310 samples, 29903 variants)
consisting of 37 fields and 391 MiB of storage, 
of which 280MiB is used by alignments ($\sim 81$X smaller 
than gzipped FASTA). The chunk size is 100 variants $\times$
10,000 samples.
Shown are the top 9 fields consuming at least 1\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
Field names (derived from source TSV file) are shortened for 
display purposes by replacing ``Genbank'' with ``G''
and ``Viridian'' with ``V''.
\label{tab-sc2-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
\bottomrule
\end{tabular}
\end{table}

@@

Decompressed, the Viridian dataset consists of 48
FASTA files totalling 125 GiB of storage and a 1.4 GiB
TSV file for sample metadata.
When aligned to the reference and converted to Zarr, the
entire dataset is stored in a single 391 MiB Zip file.
When compressed with bgzip (so they can be indexed; see next 
paragraph) the alignments require 22 GiB of storage.
Table~\ref{tab-sc2-data} summarises the storage used by individual
arrays, and shows that the alignment data requires 298 MiB,
around 81X smaller than gzipped FASTA.
The original dataset is distributed as xz compressed FASTA files, which
achieves even higher compression levels at a total of 218 MiB.
However, these files must be decompressed in order to access the data
using standard FASTA processing techniques.

@@

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing
high-performance analysis tools,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. 
Using the TensorStore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend 
(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats),
for which the core functionality is defined in a 264 line C++ file.
TensorStore provides efficient read and write access to array
data in several formats, used, for example, in
checkpointing~\cite{orbax-tensorstore} 
TensorFlow model training~\citep{tensorflow2015-whitepaper}.
It supports several cloud platforms, and is based on an
asynchronous, parallel processing model (see Discussion).

@@

\subsection{Future applications: GPU acceleration}
GPU acceleration has revolutionised scientific computing by leveraging 
their massively parallel architecture to achieve significant speedups in
computationally intensive tasks such as simulations, data analysis, and machine
learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
mittal2014survey,owens2008gpu}
This has had a transformative impact across many 
% TODO More
fields~\citep[e.g.][]{stone2007accelerating,pandey2022transformational}.
While early GPU programming required expertise in
low-level languages like CUDA or OpenCL, the advent of high-level Python
libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
and PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
has made GPU acceleration far more accessible.
Deep learning~\citep{lecun2015deep} has led to major breakthroughs in 
several domains in recent 
% TODO more?
years~\citep[e.g.][]{jumper2021highly,merchant2023scaling}
and depends crucially on processing large volumes of data
on GPUs or other custom hardware accelerators.
The Python data science ecosystem is particularly powerful in this area,
with GPU-native libraries like
PyTorch~\cite{paszke2019pytorch}
and TensorFlow~\citep{tensorflow2015-whitepaper}
providing the basic infrastructure for the field.

@@

To demonstrate the ease at which non-trivial, novel calculations 
can be performed on a GPU,
we first reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the central operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel. 
This is very similar to the Numba compiled function used in other
experiments, with minimal changes required to take advantage of
on-GPU parallelism. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82 MiB), decompress each to a local
NumPy array and then copy the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with a
data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460 MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer time to 138ms and processing time to 6ms.
